Tsbrowse: an interactive browser for Ancestral Recombination Graphs

### 2025-06-10 manuscript/revision/response_to_editor.tex

\documentclass{letter}
\usepackage{xcolor}
\usepackage{hyperref}
\signature{Savita Karthikeyan}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{}

\opening{Dear Alison,}

We are delighted that the reviewers have found our manuscript suitable for publication 
in Bionformatics. We thank you, the editors and all the reviewers for the careful 
consideration of our manuscript, and value the kind words and constructive feedback 
provided on our work. We have now revised the manuscript based on the reviewers' 
suggestions. Below is our point-by-point response to the 
reviewers' comments in blue. We have incorporated as many suggestions as possible, 
and where unable to do so we have indicated the reason.\\

\hrule

Reviewer: 1

Comments to the Author \\

Summary: In this manuscript, the authors present tsbrowse, an open-source Python-based 
application for the interactive visualization of the components of Ancestral Recombination 
Graphs (ARGs). Because ARGs (in principle) contain all knowable information concerning 
coalescence and recombination in the history of a set of samples, there is great excitement 
in developing methods for inferring ARGs from empirical genomic sequence data, as well as 
for learning about evolutionary processes from patterns in inferred ARGs. A crucial component 
of both these endeavors is visualizing ARGs, which is difficult or unwieldy. Tsbrowse therefore 
represents an exciting contribution to the field which is likely to be of high utility and 
to be highly cited.

Opinion: This paper was straightforward, clear, and well written. I think it does a good job 
setting up the intellectual context for ARG-based research and advocating for the utility of 
tsbrowse. I had two minor, nitpicky comments, but overall, it was great.  Thanks for a fun and 
easy review!

1. I think the order in which the supplemental figures are cited doesn’t match their numbering 
(i.e., S4 is cited before S1)\\
\textcolor{blue}{Thanks to Reviewer 1 for pointing out this error. We have now rearranged the 
supplementary figures so that the citations appear in ascending order.}

2. In the discussion, the authors mention that methods for visualization of tree topology can 
typically only handle a few hundred nodes, but it might be worth doing a “(but see…)” for some 
of the approaches folks have taken for visualizing much larger trees (e.g., nextstrain, onezoom).\\
\textcolor{blue}{We agree with the reviewer-  it is important to acknowledge previous/on-going visualisation efforts 
for larger trees. We had already included a citation to the OneZoom project (and the Treenome browser) in the 
sentence that followed the one in question "… Visualisation of large-scale tree topologies with millions of nodes 
requires much more sophisticated approaches to capture topological features at different scales, 
and is an active research area". We are thankful to the reviewer for bringing Nextstrain to our 
attention, and have now included Hadfield et al. as a citation.}

Reviewer: 2\\

Comments to the Author\\

I have only a few minor comments:\\
1. "and storing as a compressed .tsbrowse file. (The .tsbrowse file is also a valid input for
tszip, a general utility for compressing tskit files.)" Sorry, but I don't understand this sentence. 
More precisely, both the compressed and the uncompressed files have the .tbrowse extension? If yes, 
that's a weird choice and I would like to see some justification. You should also make clear if tszip 
is a part of tskit or of twbrowse.\\
\textcolor{blue}{To clarify, the tszip is part of the tskit ecosystem. It is not part of the tsbrowse 
package but can be used to compress the tables output from the tsbrowse preprocess step. Thanks to 
the reviewer for this pointing out this potentially unclear statement. We have now deleted the text 
in brackets, which we believe isn't crucial for understanding the data model.}

2. "web-browser" - web browser\\
\textcolor{blue}{Fixed, thanks for pointing out the typographical error.}

3. "Fig S4, for example, shows a screenshot of tsbrowse". This is a matter of personal taste, but I do
not like to refer to a Figure in the supplementary materia just as if it were in the article; if the 
Figure is relevant, it should be in the article. If it's not relevant, that you should refer to it without
hinting that it is important.\\
\textcolor{blue}{We agree with the reviewer's opinion, and believe that Figure S4 is crucial 
in order to demonstrate the application of tsbrowse to large-scale datasets (which is a key design 
feature of the app). However, we are unable to move this figure to the main matter considering journal 
guidelines for page limit.}

4. Looking at the github repo, I can find the installation instruction using pip only. In my experience 
pip is quite picky. I strongly suggest you to provide tsbrowse via conda-forge/bio-conda or a containerized 
solution (e.g. docker).\\
\textcolor{blue}{Thanks to the reviewer for the suggestion to provide alternate installation solutions. 
We have now created a Docker container in order to ensure a smooth experience across computing environments. 
Links have been updated in the manuscript and app documentation}
\textcolor{red}{[TO DO; after release of new package]}

Reviewer: 3

Comments to the Author
This paper offers a browser-based visualisation tool for ancestral recombination graphs (ARGs), which is 
a data structure representing gene genealogies for a sample of DNA sequences. The complication in the p
resence of recombination is that these structures are not simple trees, but - because of the possibility 
of recombination - graphs with reticulated nodes, which makes exploratory data analysis difficult.

I agree with the authors that there is currently a lack of visualisation tools for ARGs, and I expect the 
current paper to provide a useful addition to the catalog of tools for ARG simulation and inference. Tsbrowse 
integrates with the widely used tskit ecosystem of software tools, developed by some of the current authors, 
which will facilitate a rapid uptake. The authors make a compelling argument for the need for visualisation 
tools - Fig S1 in particular shows that four different and popular tools for ARG inference are giving 
qualitatively completely different insights into their reconstructed ARGs (incidentally, it's a shame that 
that figure couldn't make it into the main paper, though I recognise that this journal has strict page limits).\\
\textcolor{blue}{We appreciate the reviewer's feedback on the utility of tsbrowse, and acknowledge the 
importance of S1 in highlighting the need for visualisation. However, considering the necessity of the 
existing main display items to establish the fundamental aspects of tsbrowse, we have decided to retain S1 
in the supplementary. We agree that S1 would have been a compelling main display item had the page/figure 
limits of the journal allowed for it.}

My main comment is that tsbrowse does not actually seem to be an ARG visualisation tool. It is a useful way 
to visualise mutations, tree-sequence edges, and summary statistics, but nowhere will you actually see a tree 
(much less an ARG). On rereading the paper I see that the manuscript has been carefully worded not to include 
such a claim, but a casual reader might expect tsbrowse to be able to do so, as other tools such as IcyTree can. 
I do get fed up with reviewers tossing additional requests on top of an already polished piece of work, but 
in this case I think it would be a helpful addition! - would it really be too much work for tsbrowse to be 
able to draw at least the local trees? (For researchers with smaller datasets, being able to interrogate the 
fine details of an ARG would be equally valuable as the complementary selling-points around scalability.) At 
the very least I would like to see the manuscript at least be a little more up-front about this omission.\\

\textcolor{blue}{We thank the reviewer for making this important observation, and agree that drawing 
(parts of) local trees would be a great feature. For smaller datasets, tools such as arg-visualiser 
(\url{https://pypi.org/project/tskit-arg-visualizer/}) and existing tskit infrastructure, for example 
the \texttt{draw\_text()} and \texttt{draw\_svg()} already provide this functionality. A key 
motivation for this work is to guide large-scale inference, and implementing tree visualisation that is interpretable 
at this scale would require significant thought towards design as well as considerable effort. We therefore 
think that implementing this feature is not feasible for this work and is better scoped as future work or as 
a separate endeavour. In the text, we do state that tsbrowse is a visualiser for the information structure 
in ARGs, which we hope implies that it is a visualiser for the properties of the ARG, rather than the ARG 
topology itself. We have also already explicitly stated in the Discussion that drawing local trees is an 
important future direction.}\\

Minor comments:\\

Manuscript p1: 'it is only with recent breakthroughs in inference methods...that practical application 
has become possible'. I appreciate the need to emphasise the recent increase in interest in inference for ARGs, 
but I think this is selling older methods a little short. There are ARG inference methods going back at least 
as far as Griffiths and Marjoram (1996) 'Ancestral Inference from Samples of DNA Sequences' JCB 3(4):479-502. 
True they are not scalable in the manner of recent methods, but they *are* practical applications, and they have 
the advantage that they actually sample from the claimed model without introducing any model heuristics.
\textcolor{blue}{Thanks to the reviewer for their feedback on this statement on early methods. We agree, 
and have revised the text to acknowledge the contributions of foundational work in ARG inference.}

Manuscript p1: Typo 'developements'.\\
\textcolor{blue}{Thanks, this is now fixed.}

Manuscript p2: Where you mention 'Bokeh', can you clarify what this is (e.g. with a citation)?\\
\textcolor{blue}{Yes, Bokeh is a Python library for interactive visualisation and a citation to it 
has been added.}
\textcolor{red}{TO FIX! Citation appears as a question mark!!.}

Supplement p1, Sec 1.4: "The ARG shown in Fig 4" should say "The ARG shown in Suppl. Fig 4" ?\\
\textcolor{blue}{Thanks, this is now fixed.}\\
\textcolor{red}{(TO DO: make sure to replace links in suppl pdf!)}

Fig S2 caption - 'Consortium (2015)' is citing incorrectly.\\
\textcolor{blue}{Fixed (removed this citation as it was not required)}

From the online documentation, Node page: \url{https://tskit.dev/tsbrowse/docs/stable/nodes.html}
"The histograms at the bottom show the distributions of node spans over different dimensions. 
The leftmost histogram summarises the span of nodes on the sequence; the middle plot summarises 
the span of nodes over time and the rightmost plot summarises the edge “area” defined as the product 
of sequence span and time span for each node."

I didn't understand this. The 'nodes' section of the visualiser only seems to provide one histogram, 
not three.\\

\textcolor{blue}{Thanks to the reviewer for pointing out this documentation error. The Nodes section 
does indeed have one histogram and the documentation has been updated to reflect this.}

Installation notes:

I had a few problems getting tsbrowse to work. There are two different sets of installation instructions:

1. https://pypi.org/project/tsbrowse/
2. https://tskit.dev/tsbrowse/docs/stable/contributing.html

I use a Mac+Homebrew+Chrome and Option 2 seems more fitting. ("python -m pip install tsbrowse" as in 
option 1 will fail because Homebrew's Python environment is externally managed.)

\textcolor{blue}{We apologise for any confusion on installation instructions. The recommended way to install 
tsbrowse is via pip within a virtual environment. We have fixed the documentation to show only one set of 
instructions (i.e, for the user) at the top.}

Bug(?):

I frequently had problems rendering the window in the 'Mutations' tab, which tended to get stuck at the 
circling wheel. When clicking the 'Mutations' button, my terminal returned the following error:

2025-05-27 12:29:41,646 [5441] ERROR    tornado.application: Exception in callback functools.partial(>, .wrapped() 
done, defined at \\
/Users/reviewer/myenv/lib/python3.13/site-packages/panel/io/server.py:153> 
exception=AbbreviatedException(, ValueError('Supplied cmap winter not found among matplotlib, bokeh, or colorcet 
colormaps.'), )>)

\textcolor{blue}{Thanks very much to the reviewer for bringing this bug to our attention. It appears to be a 
packaging error. An issue has been raised on the tsbrowse code repository 
\url{https://github.com/tskit-dev/tsbrowse/issues/248}. We are working on a solution, and will release a new 
Python package as soon as this is fixed.}\\

\hrule

We have also addressed the editorial comments and ensured the following:
\begin{itemize}
    \item \textcolor{red}{TO DO.} Code has been published on Zenodo and archival DOI updated in the code availability in addition 
        to code repository
    \item \textcolor{red}{TO DO.} Manuscript adheres to mandatory template format: \textcolor{red}{INSERT TEMPLATE LINK}
    \item \textcolor{red}{TO DO.} Supplementary material attached as separate pdf to main matter
    \item \textcolor{red}{TO DO.} Following .tex files included: .tex, .cls, .bib, .bst and .ps 
    \item \textcolor{red}{TO DO.} Mark up of changes in the revised text included as \textcolor{red}{INSERT NAME OF DOC}
    \item \textcolor{red}{TO DO.} Conflicts of interests declared for all authors
\end{itemize}

We look forward to hearing from you about our revised manuscript. Please let us know if you require any 
further clarification.

On behalf of all tsbrowse authors,
\closing{Sincerely,}

\end{letter}
\end{document}

### 2025-06-10 manuscript/paper.tex
typo in repo link
minor revisions and response to reviewers

\abstract{
    %\textbf{Motivation:}
    Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
    ancestry for a set of recombining sequences. The ability to capture the
    evolutionary history of samples makes ARGs valuable in a wide range of
    applications in population and statistical genetics. ARG-based approaches are
    increasingly becoming a part of genetic data analysis pipelines due to
    breakthroughs enabling ARG inference at biobank-scale. However, there is a lack
    of visualisation tools, which are crucial for validating inferences and
    generating hypotheses.
    %\textbf{Results:} 
    We present \texttt{tsbrowse}, an open-source Python web-app for the interactive
    visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges
    and mutations. We demonstrate the application of \texttt{tsbrowse} to various
    data sources and scenarios, and highlight its key features of browsability
    along the genome, user interactivity, and scalability to very large sample
    sizes.\\ \textbf{Availability:}\\ Python package:
    \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version:
    \texttt{\url{https://github.com/tskit.dev/tsbrowse}}, \\ Documentation:
    \texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
    %\textbf{Contact:} \href{name@email.com}{name@email.com}\\
    %\textbf{Supplementary information:} Available at
    %\textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics}
    %online}.}}
}

@@

\section{Introduction}
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a
recombining
species, and are currently the subject of intense research
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
    wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
    speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
    gunnarsson2024scalable,deng2024robust} that practical application has
become possible. Varied applications have been recently proposed,
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2025geographic},
more powerful approaches to quantifying genetic
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree}.
While these developements are exciting, the performance 
of these new methods depends critically on the accuracy of the inferred ARGs.
Although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data are 
understudied. In particular, there is a critical lack of software 
infrastructure to support evaluation and quality control of 
inferred ARGs.

@@

\subsection{Architecture}
\texttt{Tsbrowse} is a modular client-server application
written in Python, optimised for ARGs with millions
of samples (Fig.~\ref{fig:Figure_1}).
The application has two basic commands: \texttt{preprocess}
and \texttt{serve}. The \texttt{preprocess} command
takes an input ARG file and augments the \texttt{tskit}
tables with additional columns,
precomputing all the information required for visualisation,
and storing as a compressed \texttt{.tsbrowse} file.
(The \texttt{.tsbrowse} file is also a valid input for
\texttt{tszip}, a general utility for compressing \texttt{tskit}
files.)
To visualise an ARG we then run \texttt{tsbrowse serve}, which
by default will open a browser window on the local machine, but
also supports running the server on a remote machine across the
network.
This client-server architecture has some important advantages over
a monolithic single-machine approach. Most importantly,
the ARG being visualised remains in-situ and does not need to be
downloaded from the server.

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interactively navigates the
ARG.
This approach allows us to summarise very large ARGs interactively;
Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of
\texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
ARG with around 2.7 million nodes and edges.

@@

% and renders visualisations. This step follows a client-server architecture. On
% the server side (the machine on which \texttt{tsbrowse serve} is run), the
% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
% visualisations based on the pre-processed data. This choice of architecture is
% prompted by the need to represent properties of large-scale ARGs in a
% meaningful and performant way. For example, a visual representation of millions
% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
% lead to performance bottlenecks and overplotting. Datashader
% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the

% \texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
% which makes it convenient to handle large datasets. 
% The initial step (\texttt{tsbrowse
% preprocess}) prepares the data for visualisation. It can be performed
% separately on a large machine, and the output stored for later use. The second
% step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
% and renders visualisations. This step follows a client-server architecture. On
% the server side (the machine on which \texttt{tsbrowse serve} is run), the
% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
% visualisations based on the pre-processed data. This choice of architecture is
% prompted by the need to represent properties of large-scale ARGs in a
% meaningful and performant way. For example, a visual representation of millions
% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
% lead to performance bottlenecks and overplotting. Datashader
% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
% server to render and interact with these visualisations. Heavy data processing
% is limited to the server side, keeping the client-side interface lightweight.
% For large datasets with complex visualisations, this achieves a smoother user
% experience.

\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
    \centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{Screenshot of the Edges tab in \texttt{tsbrowse}.
        Visualisation of the edges in a simulated ARG with a strong selective
        sweep in the middle of the genome (see Supplement for details).
        Edges are shown in the main browser pane on the left, with additional
        histograms summarising the edges on the right.
        The effect of the sweep can be seen by the lack of edges crossing the
        centre of the simulated chromosome due to an excess of recent coalescent
        events. Moving away from the focal site, the oldest edges are the
        first to rejoin the ARG, followed by more recent edges, resulting in a
        wedge-like pattern of missing edges in the centre.
        \label{fig:Figure_2}}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with pages to
describe various aspects of the ARG.
Figure~\ref{fig:Figure_2}
demonstrates the user interface using the Edges page as an example.
The plot on the left is an overview of the 33,929 edges
in a simulated ARG with a strong selective sweep (see
Sec.~\ref{subsec:sweep_simulation} for details).
Each edge is depicted as a horizontal line connecting the genomic coordinates
of the
parent and child nodes on the x-axis, and the y-axis shows the time of
either the parent or the child, as chosen by the user.
The user can interact with the main ``genome browser'' window
using a set of controls on the top-right corner (provided by Bokeh),
allowing the user to pan and zoom as required. The histograms to the
right then summarise the edges as depicted in the browser window.
A similar browser interface is provided for
nodes
(e.g. Fig~\ref{fig:Supplementary_Figure_3})
and mutations
(e.g. Fig~\ref{fig:Supplementary_Figure_4}).
\texttt{Tsbrowse} also provides an interactive table viewer with flexible
search and sorting utilities, which is a valuable debugging utility for
developers.

@@

\subsection{Applications}
The purpose of \texttt{tsbrowse} is to provide an interactive view of the
\texttt{tskit} ARG data model to guide intuition, improve inference quality
control and facilitate debugging. An example of how we can deepen our understanding
using the genome-browser-like perspective provided by \texttt{tsbrowse}
is given in the simulated ARG of Fig~\ref{fig:Figure_2}, where
the gap in the edges view corresponds to the characteristic dip in diversity
of a selective sweep. Comparing this ground truth to the ARGs inferred
by four different inference methods in Fig~\ref{fig:Supplementary_Figure_1},
we can see that there are substantial qualitative differences between the
results. These differences in ancestral haplotypes illustrated by
the edges view are unlikely to be captured by the tree-by-tree distance
metrics usually used to evaluate
inference methods~\citep[e.g.][]{kelleher2019inferring,zhang2023biobank},
providing further motivation for new and improved ways to compare
simulated and inferred ARGs~\citep{fritze2024forest}.

>>>>

\abstract{
    %\textbf{Motivation:}
    Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
    ancestry for a set of recombining sequences. The ability to capture the
    evolutionary history of samples makes ARGs valuable in a wide range of
    applications in population and statistical genetics. ARG-based approaches are
    increasingly becoming a part of genetic data analysis pipelines due to
    breakthroughs enabling ARG inference at biobank-scale. However, there is a lack
    of visualisation tools, which are crucial for validating inferences and
    generating hypotheses.
    %\textbf{Results:} 
    We present \texttt{tsbrowse}, an open-source Python web-app for the interactive
    visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges
    and mutations. We demonstrate the application of \texttt{tsbrowse} to various
    data sources and scenarios, and highlight its key features of browsability
    along the genome, user interactivity, and scalability to very large sample
    sizes.\\ \textbf{Availability:}\\ Python package:
    \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version:
    \texttt{\url{https://github.com/tskit-dev/tsbrowse}}, \\ Documentation:
    \texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
    %\textbf{Contact:} \href{name@email.com}{name@email.com}\\
    %\textbf{Supplementary information:} Available at
    %\textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics}
    %online}.}}
}

@@

\section{Introduction}
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a
recombining
species, and are currently the subject of intense research
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}.
%Version 1
%While early methods laid the groundwork for practical applications, 
%it is with recent breakthroughs in inference methods ~\citep{rasmussen2014genome,
%speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
%gunnarsson2024scalable,deng2024robust} that scalability was achieved,
%enabling the widespread application of ARGs to large-scale genomic data.
%Version 2
%ARGs are fundamental objects in population genetics,
%with decades of foundational work informing their study and practical
%applications~\citep{Hudson1983,Griffiths1996,Griffiths1997}.
%~\citep{Hudson1983,Griffiths1996,Griffiths1997}.Recent breakthroughs 
%in inference methods~\citep{rasmussen2014genome, speidel2019method, 
%kelleher2019inferring, wohns2022unified, zhang2023biobank, 
%gunnarsson2024scalable, deng2024robust} have enabled their 
%widespread application to large-scale genomic data.
%Version 3
While decades of foundational work have informed the study of ARGs 
and established them as fundamental objects in population genetics
~\citep{Hudson1983,Griffiths1996,Griffiths1997}, recent breakthroughs 
in inference methods~\citep{rasmussen2014genome, speidel2019method, 
kelleher2019inferring, wohns2022unified, zhang2023biobank, 
gunnarsson2024scalable, deng2024robust} have expanded their application
to large-scale, genome-wide data.

Varied applications have been proposed,
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2025geographic},
more powerful approaches to quantifying genetic
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree}.
While these developments are exciting, the performance 
of these new methods depends critically on the accuracy of the inferred ARGs.
Although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data are 
understudied. In particular, there is a critical lack of software 
infrastructure to support evaluation and quality control of 
inferred ARGs.

@@

\subsection{Architecture}
\texttt{Tsbrowse} is a modular client-server application
written in Python, optimised for ARGs with millions
of samples (Fig.~\ref{fig:Figure_1}).
The application has two basic commands: \texttt{preprocess}
and \texttt{serve}. The \texttt{preprocess} command
takes an input ARG file and augments the \texttt{tskit}
tables with additional columns,
precomputing all the information required for visualisation,
and storing as a compressed \texttt{.tsbrowse} file.
%(The \texttt{.tsbrowse} file is also a valid input for
%\texttt{tszip}, a general utility for compressing \texttt{tskit}
%files.)
To visualise an ARG we then run \texttt{tsbrowse serve}, which
by default will open a browser window on the local machine, but
also supports running the server on a remote machine across the
network.
This client-server architecture has some important advantages over
a monolithic single-machine approach. Most importantly,
the ARG being visualised remains in-situ and does not need to be
downloaded from the server.

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web browser
for display, and dynamically updates as the user interactively navigates the
ARG.
This approach allows us to summarise very large ARGs interactively;
Fig~\ref{fig:Supplementary_Figure_1}, for example, shows a screenshot of
\texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
ARG with around 2.7 million nodes and edges.

@@

% and renders visualisations. This step follows a client-server architecture. On
% the server side (the machine on which \texttt{tsbrowse serve} is run), the
% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
% visualisations based on the pre-processed data. This choice of architecture is
% prompted by the need to represent properties of large-scale ARGs in a
% meaningful and performant way. For example, a visual representation of millions
% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
% lead to performance bottlenecks and overplotting. Datashader
% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_1} for a
% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the

% \texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
% which makes it convenient to handle large datasets. 
% The initial step (\texttt{tsbrowse
% preprocess}) prepares the data for visualisation. It can be performed
% separately on a large machine, and the output stored for later use. The second
% step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
% and renders visualisations. This step follows a client-server architecture. On
% the server side (the machine on which \texttt{tsbrowse serve} is run), the
% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
% visualisations based on the pre-processed data. This choice of architecture is
% prompted by the need to represent properties of large-scale ARGs in a
% meaningful and performant way. For example, a visual representation of millions
% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
% lead to performance bottlenecks and overplotting. Datashader
% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_1} for a
% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
% server to render and interact with these visualisations. Heavy data processing
% is limited to the server side, keeping the client-side interface lightweight.
% For large datasets with complex visualisations, this achieves a smoother user
% experience.

\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
    \centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{Screenshot of the Edges tab in \texttt{tsbrowse}.
        Visualisation of the edges in a simulated ARG with a strong selective
        sweep in the middle of the genome (see Supplement for details).
        Edges are shown in the main browser pane on the left, with additional
        histograms summarising the edges on the right.
        The effect of the sweep can be seen by the lack of edges crossing the
        centre of the simulated chromosome due to an excess of recent coalescent
        events. Moving away from the focal site, the oldest edges are the
        first to rejoin the ARG, followed by more recent edges, resulting in a
        wedge-like pattern of missing edges in the centre.
        \label{fig:Figure_2}}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with pages to
describe various aspects of the ARG.
Figure~\ref{fig:Figure_2}
demonstrates the user interface using the Edges page as an example.
The plot on the left is an overview of the 33,929 edges
in a simulated ARG with a strong selective sweep (see
Sec.~\ref{subsec:sweep_simulation} for details).
Each edge is depicted as a horizontal line connecting the genomic coordinates
of the
parent and child nodes on the x-axis, and the y-axis shows the time of
either the parent or the child, as chosen by the user.
The user can interact with the main ``genome browser'' window
using a set of controls on the top-right corner (provided by 
Bokeh~\citep{bokeh}),allowing them to pan and zoom as required. 
The histograms to the
right then summarise the edges as depicted in the browser window.
A similar browser interface is provided for
mutations
(e.g. Fig~\ref{fig:Supplementary_Figure_1})
and nodes
(e.g. Fig~\ref{fig:Supplementary_Figure_2}).
\texttt{Tsbrowse} also provides an interactive table viewer with flexible
searching and sorting utilities, which is a valuable debugging utility for
developers.

@@

\subsection{Applications}
The purpose of \texttt{tsbrowse} is to provide an interactive view of the
\texttt{tskit} ARG data model to guide intuition, improve inference quality
control and facilitate debugging. An example of how we can deepen our understanding
using the genome-browser-like perspective provided by \texttt{tsbrowse}
is given in the simulated ARG of Fig~\ref{fig:Figure_2}, where
the gap in the edges view corresponds to the characteristic dip in diversity
of a selective sweep. Comparing this ground truth to the ARGs inferred
by four different inference methods in Fig~\ref{fig:Supplementary_Figure_3},
we can see that there are substantial qualitative differences between the
results. These differences in ancestral haplotypes illustrated by
the edges view are unlikely to be captured by the tree-by-tree distance
metrics usually used to evaluate
inference methods~\citep[e.g.][]{kelleher2019inferring,zhang2023biobank},
providing further motivation for new and improved ways to compare
simulated and inferred ARGs~\citep{fritze2024forest}.

Interest in ARG-based methods is burgeoning, but the methods are new
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensable
element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_4} 
shows a region of the 1000G data with gaps in site density that are spanned by
 exceptionally long edges, which are likely to bias downstream statistics.
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm.
Fig~\ref{fig:Supplementary_Figure_2} shows the genomic spans of
ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear
excess of long haplotypes in the very ancient past. These insights
have helped guide development and
may lead to significant improvements in performance.
